Testing Quasar Unification: I. Radiative Transfer in Clumpy Winds

### 2015-12-22 paper_mnras.tex
some more late changes

\maketitle
\begin{abstract} 
Various unification schemes interpret the complex phenomenology of quasars and luminous active galactic nuclei (AGN) in terms of a simple picture involving a central black hole, an accretion disc and an associated outflow. Here, we continue our tests of this paradigm by comparing quasar spectra to synthetic spectra of biconical disc wind models, produced with our state-of-the-art Monte Carlo radiative transfer code. Previously, we have shown that we could produce synthetic spectra resembling those of observed broad absorption line (BAL) quasars, but only if the X-ray luminosity was limited to about 
10$^{43}$~erg~s$^{-1}$. Here, we show that clumping the outflow maintains the ionization state necessary for BAL features in the rest-frame UV at more realistic X-ray luminosities. Our fiducial model shows good agreement with X-ray properties of AGN and the dense, X-ray heated wind produces strong line emission in, e.g., \la\ and \civline at low inclinations. At high inclinations, the synthetic spectra possess prominent LoBAL features. Despite these successes, we cannot reproduce all the emission lines seen in quasar spectra with the correct equivalent width ratios, and we find an angular dependence of emission line equivalent width despite the remarkable similarities in the observed spectral properties of BAL and non-BAL quasars. Overall, our work suggests that a biconical wind can reproduce much of the qualitative behaviour expected from a unified model, but we cannot yet provide quantitative matches with quasar properties at all viewing angles. Whether disc winds can successfully unify quasars is therefore still an open question.
\end{abstract}

\begin{keywords}
galaxies: active; accretion, accretion discs; (galaxies:) quasars: emission lines;
(galaxies:) quasars: absorption lines; methods: numerical; radiative transfer.
\end{keywords}

@@

At high inclinations, the synthetic spectra exhibit blue-shifted BALs in \al\ and \mg --
the absorption lines seen in LoBALQSOs, and we even see absorption in Fe~\textsc{ii}
at the highest inclinations. Line profiles in velocity space 
for \civ, \al\ and \mg, are shown in figure~\ref{fig:lobal} for a range
of BALQSO viewing angles. We find that ionization stratification
of the wind causes lower ionization material to have a smaller covering factor, 
as demonstrated by figures~\ref{fig:wind} and \ref{fig:lobal}.
This confirms the behaviour expected from a unification model such as Elvis (2000). 
LoBALs are only present at viewing angles close to edge-on ($i>75^\circ$),
as predicted by polarisation results \citep{brotherton1997}.
As observed in a BALQSO sample by \cite{filizak2014}, we find that
BAL troughs are wider and deeper when low ionization absorption features are present,
and high ionization lines have higher blue-edge velocities than the 
low ionization species.
There is also a correlation between the strength of LoBAL features
and the amount of continuum attenuation at that sightline, particularly
blueward of the Lyman edge as the low ionization base 
intersects the line-of-sight. 
A model such as this therefore predicts that LoBALQSOs and FeLoBALQSOs 
have stronger Lyman edge absorption and 
are more Compton-thick than HiBALQSOs and Type 1 quasars.
An edge-on scenario also offers a potential explanation for the rarity of LoBAL and
FeLoBAL quasars, due to a foreshortened and attenuated continuum, 
although BAL fraction inferences are fraught with complex selection 
effects \citep{goodrich1997,krolikvoit1998}.

@@

We have carried out MCRT simulations using a simple
prescription for a biconical disc wind, with
the aim of expanding on the work of H13. To do this, we introduced two main
improvements: First, we included a simple treatment of clumping, and second, 
we improved the modelling of recombination lines by treating H and He as
`macro-atoms'. Having selected a fiducial model from an initial simulation grid,
we assessed the viability of such a model for geometric 
unification of quasars, and found the following main points:
\begin{enumerate}
\item Clumping moderates the ionization state
sufficiently to allow for the 
formation of strong UV BALs while agreeing well with the X-ray
properties of luminous AGN and quasars. 
\smallskip
\item A clumpy outflow model naturally 
reproduces the range of ionization states
expected in quasars, due to it's stratified density
and temperature structure. 
LoBAL line profiles are seen at a subset of viewing angles, and Fe~\textsc{ii}
absorption is seen at particularly high inclinations. 
\smallskip
\item The synthetic spectra show \la\ line and weak He~\textsc{ii}~$1640$\AA\ line
as a result of our improved treatment of recombination using macro-atoms. We also see
Balmer emission lines and a Balmer recombination continuum in the optical spectrum, but this
is only really significant at high inclination where to continuum is suppressed.  
\smallskip
\item The higher X-ray luminosity causes a significant 
increase in the strength of the collisionally excited emission
lines produced by the model. 
However, the equivalent width ratios of the emission lines do not match
observations, suggesting that a greater volume of dense ($n_e\sim10^{10}$~cm$^{-3}$)
material may be required.
\smallskip
\item The line EWs in the synthetic spectra increase with inclination.
BAL and non-BAL quasar composites have comparable EWs, so our model
fails to reproduce this behaviour.
 If the BLR emits fairly isotropically then for a 
foreshortened, limb-darkened accretion disc 
it is not possible to achieve line ratios at low inclinations 
that are comparable to those at high inclinations. 
We suggest that understanding the angular distribution of 
line and continuum emission is a crucial question for theoretical models.
% Furthermore, this is conclusion is independent of the assumed 
% BLR geometry and size.
\end{enumerate}
Our work confirms a number of expected outcomes from a geometric unification 
model, and suggests that a simple biconical geometry such as this can come close to 
explaining much of the  phenomenology of quasars. However, our conclusions pose 
some challenges to a picture in which BALQSOs are
explained by an {\em equatorial} wind rising from a classical thin disc, and suggest 
the angular distribution of emission is important to understand if this 
geometry is to be refuted or confirmed. We suggest that obtaining reliable 
observational orientation indicators and 
exploring a wider parameter space of outflow geometries in simulations
are obvious avenues for future work.

>>>>

\maketitle
\begin{abstract} 
Various unification schemes interpret the complex phenomenology of quasars and luminous active galactic nuclei (AGN) in terms of a simple picture involving a central black hole, an accretion disc and an associated outflow. Here, we continue our tests of this paradigm by comparing quasar spectra to synthetic spectra of biconical disc wind models, produced with our state-of-the-art Monte Carlo radiative transfer code. Previously, we have shown that we could produce synthetic spectra resembling those of observed broad absorption line (BAL) quasars, but only if the X-ray luminosity was limited to about 
10$^{43}$~erg~s$^{-1}$. Here, we show that clumping the outflow can maintain the ionization state necessary for BAL features in the rest-frame UV at more realistic X-ray luminosities. Our fiducial model shows good agreement with X-ray properties of AGN and the dense, X-ray heated wind produces strong line emission in, e.g., \la\ and \civline at low inclinations. At high inclinations, the synthetic spectra possess prominent LoBAL features. Despite these successes, we cannot reproduce all the emission lines seen in quasar spectra with the correct equivalent-width ratios, and and we find an angular dependence of emission-line equivalent width despite the remarkable similarities in the observed emission line properties of BAL and non-BAL quasars. Overall, our work suggests that a biconical wind can reproduce much of the qualitative behaviour expected from a unified model, but we cannot yet provide quantitative matches with quasar properties at all viewing angles. Whether disc winds can successfully unify quasars is therefore still an open question.
\end{abstract}

\begin{keywords}
galaxies: active -- accretion, accretion discs -- (galaxies:) quasars: emission lines --
(galaxies:) quasars: absorption lines -- methods: numerical -- radiative transfer.
\end{keywords}

@@

At high inclinations, the synthetic spectra exhibit blue-shifted BALs in \al\ and \mg --
the absorption lines seen in LoBALQSOs, and we even see absorption in Fe~\textsc{ii}
at the highest inclinations. Line profiles in velocity space 
for \civ, \al\ and \mg, are shown in Fig.~\ref{fig:lobal} for a range
of BALQSO viewing angles. We find that ionization stratification
of the wind causes lower ionization material to have a smaller covering factor, 
as demonstrated by figures~\ref{fig:wind} and \ref{fig:lobal}.
This confirms the behaviour expected from a unification model such as Elvis (2000). 
LoBALs are only present at viewing angles close to edge-on ($i>75^\circ$),
as predicted by polarisation results \citep{brotherton1997}.
As observed in a BALQSO sample by \cite{filizak2014}, we find that
BAL troughs are wider and deeper when low ionization absorption features are present,
and high ionization lines have higher blue-edge velocities than the 
low ionization species.
There is also a correlation between the strength of LoBAL features
and the amount of continuum attenuation at that sightline, particularly
blueward of the Lyman edge as the low ionization base 
intersects the line-of-sight. 
A model such as this therefore predicts that LoBALQSOs and FeLoBALQSOs 
have stronger Lyman edge absorption and 
are more Compton-thick than HiBALQSOs and Type 1 quasars.
An edge-on scenario also offers a potential explanation for the rarity of LoBAL and
FeLoBAL quasars, due to a foreshortened and attenuated continuum, 
although BAL fraction inferences are fraught with complex selection 
effects \citep{goodrich1997,krolikvoit1998}.

@@

We have carried out MCRT simulations using a simple
prescription for a biconical disc wind, with
the aim of expanding on the work of H13. To do this, we introduced two main
improvements: First, we included a simple treatment of clumping, and second, 
we improved the modelling of recombination lines by treating H and He as
`macro-atoms'. Having selected a fiducial model from an initial simulation grid,
we assessed the viability of such a model for geometric 
unification of quasars, and found the following main points:
\begin{enumerate}
\item Clumping moderates the ionization state
sufficiently to allow for the 
formation of strong UV BALs while agreeing well with the X-ray
properties of luminous AGN and quasars. 
\smallskip
\item A clumpy outflow model naturally 
reproduces the range of ionization states
expected in quasars, due to its stratified density
and temperature structure. 
LoBAL line profiles are seen at a subset of viewing angles, and Fe~\textsc{ii}
absorption is seen at particularly high inclinations. 
\smallskip
\item The synthetic spectra show \la\ line and weak He~\textsc{ii}~$1640$\AA\ line
as a result of our improved treatment of recombination using macro-atoms. We also see
Balmer emission lines and a Balmer recombination continuum in the optical spectrum, but this
is only really significant at high inclination where to continuum is suppressed.  
\smallskip
\item The higher X-ray luminosity causes a significant 
increase in the strength of the collisionally excited emission
lines produced by the model. 
However, the equivalent-width ratios of the emission lines do not match
observations, suggesting that a greater volume of dense ($n_e\sim10^{10}$~cm$^{-3}$)
material may be required.
\smallskip
\item The line EWs in the synthetic spectra increase with inclination.
BAL and non-BAL quasar composites have comparable EWs, so our model
fails to reproduce this behaviour.
 If the BLR emits fairly isotropically then for a 
foreshortened, limb-darkened accretion disc 
it is not possible to achieve line ratios at low inclinations 
that are comparable to those at high inclinations. 
We suggest that understanding the angular distribution of 
line and continuum emission is a crucial question for theoretical models.
% Furthermore, this is conclusion is independent of the assumed 
% BLR geometry and size.
\end{enumerate}
Our work confirms a number of expected outcomes from a geometric unification 
model, and suggests that a simple biconical geometry such as this can come close to 
explaining much of the  phenomenology of quasars. However, our conclusions pose 
some challenges to a picture in which BALQSOs are
explained by an {\em equatorial} wind rising from a classical thin disc, and suggest 
the angular distribution of emission is important to understand if this 
geometry is to be refuted or confirmed. We suggest that obtaining reliable 
observational orientation indicators and 
exploring a wider parameter space of outflow geometries in simulations
are obvious avenues for future work.
